### 2021-01-22 paper2/4-experiments.tex
Some meager progress

% \begin{figure*}[!htb]
%   \centering
%   \caption{Performance of our methods (cVAE and cFlow) on the tabular UCI Adult Dataset. This is a common benchmark in fair machine learning literature. Each model depicts the result of a ownstream classifier trained on the embeddings produced by one of the models. (Left) Accuracy of a downstream classifier. (Middle) Demographic Parity of a downstream classifier. (Right) Equality of Opportunity of a downstream classifier.}\label{fig:adult_results}
% \end{figure*}
% \subsubsection{Synthesising Dataset Bias}
\textbf{Synthesising Dataset Bias.}
\noindent For our experiments, we require a training set that exhibits a strong spurious correlation, together with a test set that does not.
For cMNIST, this is easily satisfied as we have complete control over the data generation process.
For CelebA and  UCI Adult, on the other hand,
we have to generate the split from the existing data.
To this end, we first set aside a randomly selected portion of the dataset from which to sample the biased dataset
The portion itself is then split further into two parts:
one in which $(s=-1 \land y=-1) \lor (s=+1 \land y=+1)$ holds true for all samples, call this part $\mathcal{D}_{eq}$,
and the other part, call it $\mathcal{D}_{opp}$, which contains the remaining samples.
To investigate the behaviour at different levels of correlation,
we mix these two subsets according to a mixing factor $\eta$.
For $\eta \leq \tfrac{1}{2}$, we combine (all of) $\mathcal{D}_{eq}$
with a fraction of $2\eta$ from $\mathcal{D}_{opp}$.
For $\eta > \tfrac{1}{2}$, we combine (all of) $\mathcal{D}_{opp}$
and a fraction of $2(1 -\eta)$ from $\mathcal{D}_{eq}$.
Thus, for $\eta=0$, the biased dataset is just $\mathcal{D}_{eq}$,
for $\eta=1$ it is just $\mathcal{D}_{opp}$
and for $\eta=\tfrac{1}{2}$ the biased dataset is an ordinary subset of the whole data. The test set is simply the data remaining from the initial split.

\textbf{Evaluation protocol.}
We evaluate our results in terms of accuracy and fairness.
A model that perfectly decouples its predictions from $s$ will achieve near-uniform accuracy across all biasing-levels.
For binary $s$/$y$ we quantify the fairness of a classifier's predictions using \emph{demographic parity} (DP): the  absolute difference in the probability  of a positive prediction for each sensitive group.

@@

% \subsubsection{cMNIST.}
\textbf{cMNIST.}\label{ssec:cmnist}
The coloured MNIST (cMNIST) dataset is a variant of the MNIST dataset in which the digits are coloured.
In the training set, the colours have a one-to-one correspondence with the digit class.
In the test set (and the representative set), colours are assigned randomly.
The colours are drawn from Gaussians with 10 different means.
We follow the colourisation procedure outlined by~\cite{ln2l}, with the mean colour values selected so as to be maximally dispersed.
The full list of such values can be found in Appendix~\ref{sec:color-details}.
We produce multiple variants of the cMNIST dataset corresponding to different standard deviations $\sigma$ for the colour sampling:
$\sigma \in \{0.00, 0.01, ..., 0.05 \}$.
% The images are symmetrically zero-padded to a size of $32 \times 32$ to allow for an additional downsampling stage in the cFlow model.
% No additional data augmentation is used.
% 24,000 of the 36,000 training samples were set aside for representative, the remaining 10,000 for training the downstream classifier.
% We continue to use 10,000 samples usually reserved for testing on vanilla MNIST for that same purpose.

@@

% \subsubsection{CelebA.}
\textbf{CelebA.}
To evaluate the effectiveness of our framework on real-world image data we use the CelebA dataset~\cite{liu2015faceattributes}, consisting of 202,599 celebrity images.
These images are annotated with various binary physical  attributes, including ``gender'', ``hair color'', ``young'', etc, from which we  select our sensitive and target attributes.
The images are centre cropped and resized to $64\times64$, as is standard practice.
For our experiments, we designate ``gender'' as the sensitive attribute,
and ``smiling'' and ``high cheekbones'' as target attributes.
We chose gender as the sensitive attribute as it a common sensitive attribute in the fairness literature.
For the target attributes, we chose attributes that are harder to learn than gender and which do not correlate too strongly with gender in the dataset
(``wearing lipstick'' for example being an attribute too closely correlated with gender).
% The sampling and evaluation procedure is identical to that conducted for the UCI Adult dataset.
% We evaluate the performance of all models for mixing factors ($\eta$) of value $\{0, 0.1, ..., 1\}$. 
The model is trained on the representative set (normal subset of CelebA)
and is then used to encode the artificially biased training set and the test set.
The results for the most strongly biased training set ($\eta=0$) can be found in Fig.~\ref{fig:celeba-targets}.
Our method outperforms the baselines in accuracy and fairness.

@@

To show that our method can handle multinomial, as well as binary, sensitive attributes, we also conduct experiments with $s=\textrm{hair color}$ as a ternary attribute (``Blonde'' ``Black'', ``Brown''), excluding ``Red'' because of the paucity of samples and the noisiness of their labels. The results for these experiments can be found in Appendix~\ref{ssec:multi-s}.
% Looking at the produced invariant representations,
% The failure of the cVAE to model complex interactions between $x_u$ and $s$ is brought to bear here.
% Due to the balancing between the adversarial and reconstruction losses,
% information relevant to the downstream task of predicting ``smiling'' is lost during the pre-training phase,
% leading to poor performance at all values of $\eta$ except the most extreme ones,
% for which it does reasonably well at mitigating the sampling bias (Fig.~\ref{fig:celeba_chart}).
% Meanwhile, cFlow does not suffer this drawback -- the images remain sharp
% (see Fig.~\ref{fig:cflow_celeba_recon_y} in comparison to Fig. \ref{fig:cvae_celeba_recon_y})
% and the accuracy obtained by training on them is comparable to the baseline at intermediary values -- and at the same time matches and outperforms the cVAE at $\eta = 0$ and $\eta = 1$, respectively.
% For these same values,
% incorporating the mutual information loss proposed by~\cite{ln2l} into the classifier's loss function greatly harmed performance measured by accuracy;
% the DP is zero due to all predictions being negative.

% \begin{wrapfigure}{l}{0.6\linewidth}
\begin{figure}[t]
  \centering
  \includegraphics[width=\linewidth]{./Figures/nosinn_adult_multiplot_mini_diff.pdf}
  \caption{
      Results for the \textbf{Adult} dataset.
      The $x$-axis corresponds to the difference in positive rates.
      An ideal result would occupy the \textbf{top-left}.
  }%
  \label{fig:adult-chart}
\end{figure}
% \end{wrapfigure}
%
\noindent\textbf{Results for the UCI Adult dataset.}
The UCI Adult dataset consists of census data and is commonly used to evaluate models focused on algorithmic fairness.
Following convention, we designate ``gender'' as the sensitive attribute $s$ and whether an individual's salary is \$50,000 or greater as $y$.
We show the performance of our approach in comparison to baseline approaches in Fig. \ref{fig:adult-chart}.
We evaluate the performance of all models for mixing factors ($\eta$) $0$ and $1$. 
Results shown in Fig. \ref{fig:adult-chart} show that
%while our model fails to surpass the baseline models in terms of accuracy for the balanced case (and those close to it),
we match or exceed the baseline.
%as $\eta$ moves the dataset towards a more imbalanced setting.
In terms of fairness metrics, our approach generally outperforms the baseline models for both of $\eta$.
Detailed results can be found in the Appendix~\ref{ssec:detailed-adult}.

>>>>

% \begin{figure*}[!htb]
%   \centering
%   \caption{Performance of our methods (cVAE and cFlow) on the tabular UCI Adult Dataset. This is a common benchmark in fair machine learning literature. Each model depicts the result of a ownstream classifier trained on the embeddings produced by one of the models. (Left) Accuracy of a downstream classifier. (Middle) Demographic Parity of a downstream classifier. (Right) Equality of Opportunity of a downstream classifier.}\label{fig:adult_results}
% \end{figure*}
\subsubsection{Synthesising Dataset Bias}
For our experiments, we require a training set that exhibits a strong spurious correlation, together with a test set that does not.
For cMNIST, this is easily satisfied as we have complete control over the data generation process.
For CelebA and  UCI Adult, on the other hand,
we have to generate the split from the existing data.
To this end, we first set aside a randomly selected portion of the dataset from which to sample the biased dataset
The portion itself is then split further into two parts:
one in which $(s=-1 \land y=-1) \lor (s=+1 \land y=+1)$ holds true for all samples, call this part $\mathcal{D}_{eq}$,
and the other part, call it $\mathcal{D}_{opp}$, which contains the remaining samples.
To investigate the behaviour at different levels of correlation,
we mix these two subsets according to a mixing factor $\eta$.
For $\eta \leq \tfrac{1}{2}$, we combine (all of) $\mathcal{D}_{eq}$
with a fraction of $2\eta$ from $\mathcal{D}_{opp}$.
For $\eta > \tfrac{1}{2}$, we combine (all of) $\mathcal{D}_{opp}$
and a fraction of $2(1 -\eta)$ from $\mathcal{D}_{eq}$.
Thus, for $\eta=0$, the biased dataset is just $\mathcal{D}_{eq}$,
for $\eta=1$ it is just $\mathcal{D}_{opp}$
and for $\eta=\tfrac{1}{2}$ the biased dataset is an ordinary subset of the whole data. The test set is simply the data remaining from the initial split.

\subsubsection{Evaluation protocol}
We evaluate our results in terms of accuracy and fairness.
A model that perfectly decouples its predictions from $s$ will achieve near-uniform accuracy across all biasing-levels.
For binary $s$/$y$ we quantify the fairness of a classifier's predictions using \emph{demographic parity} (DP): the  absolute difference in the probability  of a positive prediction for each sensitive group.

@@

\subsubsection{cMNIST}\label{ssec:cmnist}
The coloured MNIST (cMNIST) dataset is a variant of the MNIST dataset in which the digits are coloured.
In the training set, the colours have a one-to-one correspondence with the digit class.
In the test set (and the representative set), colours are assigned randomly.
The colours are drawn from Gaussians with 10 different means.
We follow the colourisation procedure outlined by~\cite{ln2l}, with the mean colour values selected so as to be maximally dispersed.
The full list of such values can be found in Appendix~\ref{sec:color-details}.
We produce multiple variants of the cMNIST dataset corresponding to different standard deviations $\sigma$ for the colour sampling:
$\sigma \in \{0.00, 0.01, ..., 0.05 \}$.
% The images are symmetrically zero-padded to a size of $32 \times 32$ to allow for an additional downsampling stage in the cFlow model.
% No additional data augmentation is used.
% 24,000 of the 36,000 training samples were set aside for representative, the remaining 10,000 for training the downstream classifier.
% We continue to use 10,000 samples usually reserved for testing on vanilla MNIST for that same purpose.

@@

\subsubsection{CelebA}
To evaluate the effectiveness of our framework on real-world image data we use the CelebA dataset~\cite{liu2015faceattributes}, consisting of 202,599 celebrity images.
These images are annotated with various binary physical  attributes, including ``gender'', ``hair color'', ``young'', etc, from which we  select our sensitive and target attributes.
The images are centre cropped and resized to $64\times64$, as is standard practice.
For our experiments, we designate ``gender'' as the sensitive attribute,
and ``smiling'' and ``high cheekbones'' as target attributes.
We chose gender as the sensitive attribute as it a common sensitive attribute in the fairness literature.
For the target attributes, we chose attributes that are harder to learn than gender and which do not correlate too strongly with gender in the dataset
(``wearing lipstick'' for example being an attribute too closely correlated with gender).
% The sampling and evaluation procedure is identical to that conducted for the UCI Adult dataset.
% We evaluate the performance of all models for mixing factors ($\eta$) of value $\{0, 0.1, ..., 1\}$. 
The model is trained on the representative set (normal subset of CelebA)
and is then used to encode the artificially biased training set and the test set.
The results for the most strongly biased training set ($\eta=0$) can be found in Fig.~\ref{fig:celeba-targets}.
Our method outperforms the baselines in accuracy and fairness.

@@

To show that our method can handle multinomial, as well as binary, sensitive attributes, we also conduct experiments with $s=\textrm{hair color}$ as a ternary attribute (``Blonde'', ``Black'', ``Brown''), excluding ``Red'' because of the paucity of samples and the noisiness of their labels. The results for these experiments can be found in Appendix~\ref{ssec:multi-s}.
% Looking at the produced invariant representations,
% The failure of the cVAE to model complex interactions between $x_u$ and $s$ is brought to bear here.
% Due to the balancing between the adversarial and reconstruction losses,
% information relevant to the downstream task of predicting ``smiling'' is lost during the pre-training phase,
% leading to poor performance at all values of $\eta$ except the most extreme ones,
% for which it does reasonably well at mitigating the sampling bias (Fig.~\ref{fig:celeba_chart}).
% Meanwhile, cFlow does not suffer this drawback -- the images remain sharp
% (see Fig.~\ref{fig:cflow_celeba_recon_y} in comparison to Fig. \ref{fig:cvae_celeba_recon_y})
% and the accuracy obtained by training on them is comparable to the baseline at intermediary values -- and at the same time matches and outperforms the cVAE at $\eta = 0$ and $\eta = 1$, respectively.
% For these same values,
% incorporating the mutual information loss proposed by~\cite{ln2l} into the classifier's loss function greatly harmed performance measured by accuracy;
% the DP is zero due to all predictions being negative.

% \begin{wrapfigure}{l}{0.6\linewidth}
\begin{figure}[tb]
  \centering
  \includegraphics[width=0.6\textwidth]{./Figures/nosinn_adult_multiplot_mini_diff.pdf}
  \caption{
      Results for the \textsc{Adult} dataset.
      The $x$-axis corresponds to the difference in positive rates.
      An ideal result would occupy the \textsc{top-left}.
  }%
  \label{fig:adult-chart}
\end{figure}
% \end{wrapfigure}

\subsubsection{Results for the UCI Adult dataset}
The UCI Adult dataset consists of census data and is commonly used to evaluate models focused on algorithmic fairness.
Following convention, we designate ``gender'' as the sensitive attribute $s$ and whether an individual's salary is \$50,000 or greater as $y$.
We show the performance of our approach in comparison to baseline approaches in Fig. \ref{fig:adult-chart}.
We evaluate the performance of all models for mixing factors ($\eta$) $0$ and $1$. 
Results shown in Fig. \ref{fig:adult-chart} show that
%while our model fails to surpass the baseline models in terms of accuracy for the balanced case (and those close to it),
we match or exceed the baseline.
%as $\eta$ moves the dataset towards a more imbalanced setting.
In terms of fairness metrics, our approach generally outperforms the baseline models for both of $\eta$.
Detailed results can be found in the Appendix~\ref{ssec:detailed-adult}.
